---
title: "Data Analysis Workflow Example on PBMC Data (t-SNE, UMAP, FlowSOM, MEM)"
author: "Copyright (c) 2016-2019 by Kirsten Diggins, Sierra Barone, and Jonathan Irish, All Rights Reserved; see EULA-MEM.text for MEM license information"
date: "October 2019"
output:
  html_document:
    df_print: paged
  pdf_document:
    latex_engine: xelatex
editor_options:
  chunk_output_type: inline
---

This data set contains 7 FCS (flow cytometry standard) files. Each FCS file contains single cell data for one cell subset that is a well-established, phenotypically distinct population. This is mass cytometry data for healthy human PBMC (peripheral blood mononuclear cells). The populations were expert gated following a t-SNE analysis. The first section of the code will run two dimensionality reduction tools, UMAP and t-SNE, on the data set. Next, you will run FlowSOM on the both the UMAP and t-SNE axes to cluster, or group together, the various cell populations. Finally, you will run MEM to see enrichment scores for each of the FlowSOM clusters or populations that have been expert gated. The goal of this exercise is to run several computational tools on a single cell data set to get a feel for the workflow used in the Irish lab as well as compare the various analysis methods. The method for comparison of the cell populations by automated or manual analysis is RMSD. 

```{r setup, include=FALSE}
# Time <10 sec

# Load all libraries
# If you get an error message, you will need to try re-installing packages by 
# going back to the 00_install_tools.RMD script
library(FlowSOM)
library(flowCore)
library(Biobase)
library(ggplot2)
library(hexbin)
library(viridis)
library(ggExtra)
library(RColorBrewer)
library(cytoMEM)
library(tidyverse)
library(Rtsne)
library(uwot)
library(cowplot)

cat("\n\n...'setup' finished running")
```

```{r data_preparation, warning=FALSE}
# Time <10 sec

###### CONSTANTS TO SET ########

# cofactor for arcsinh transformation
COFACTOR = 15 

# set seed for reproducible results
OVERALL_SEED = 43

# set FlowSOM target number of clusters 
CLUSTER_NUM = 10 

################################

# read files into R by setting working directory and directing R to the fcs files
setwd(paste(getwd(), "/datafiles/PBMC", sep = ""))
files <-  dir(pattern = "*.fcs")

# convert and combine data for use in downstream analysis
data <- lapply(lapply(files, read.FCS), exprs)
combined.data = as.data.frame(do.call(rbind, mapply(
  cbind, data, "cluster" = c(1:length(data)), SIMPLIFY = F)))

# choose channels with markers to use for downstream analysis and apply arcsinh transformation with a cofactor of 15
transformed.chosen.markers <- combined.data %>%
  select(contains("-"),-contains("Ir")) %>%
  mutate_all(function(x)
    asinh(x / COFACTOR))  

cat("\n\n...'data_preparation' finished running")
```

```{r run_t-SNE}
# Time ~5 min
set.seed(OVERALL_SEED)

# the line below will run t-SNE on the scaled surface markers (to see help page 
# for t-SNE, type "?Rtsne -- enter" in console) 

# you can view t-SNE progress by opening up the console below
mytSNE = Rtsne(
  transformed.chosen.markers,                        # input scaled data
  
  dims = 2,                                          # number of final 
                                                     # dimensions
  
  initial_dims = length(transformed.chosen.markers), # number of initial 
                                                     # dimensions
  
  perplexity = 30,          # perplexity (similar to # of nearest neighbors, 
                            # will scale with data sets, cannot be greater than
                            # the number of events minus 1 divided by 3)
  check_duplicates = FALSE, 
  max_iter = 1000,          # number of iterations
  verbose = TRUE
)
tsne.data = as.data.frame(mytSNE$Y)

cat("\n\n...'run_t-SNE' finished running")

```

```{r plot_t-SNE}
# Time <10 sec

# setting aspect ratio for plots
range <- apply(apply(tsne.data, 2, range), 2, diff)
graphical.ratio.tsne <- (range[1] / range[2])

# t-SNE flat dot plot and density dot plot (1 dot = 1 cell)
tsne.plot <- data.frame(x = tsne.data[, 1], y = tsne.data[, 2])

# dot plot
ggplot(tsne.plot) + coord_fixed(ratio = graphical.ratio.tsne) + 
  geom_point(aes(x = x, y = y), cex = 0.3) + labs(x = "t-SNE 1", y = "t-SNE 2", 
  title = "PBMC Data on t-SNE Axes") + 
  theme_bw() + 
  labs(caption = "Data from Digggins et al., Nat Methods 2017, 14: 275-278 \nFlow Repository: FR-FCM-ZY63") + 
  theme(panel.grid.major = element_blank(),
              panel.grid.minor = element_blank())

# density dot plot
ggplot(tsne.plot, aes(x = x, y = y)) + 
  coord_fixed(ratio = graphical.ratio.tsne)+ geom_point(size = 0.5) +geom_density_2d_filled(bins = 39) +scale_fill_manual(values = c("NA","NA","NA","NA","NA","NA","NA","NA","NA","NA","NA","NA","NA",viridis::viridis(28,option = "A"))) +
labs(x = "t-SNE 1", y = "t-SNE 2", 
  title = "Density on t-SNE Axes") + theme_bw() + 
  labs(caption = "Data from Diggins et al., Nat Methods 2017, 14: 275-278 \nFlow Repository: FR-FCM-ZY63") + 
  theme(panel.grid.major = element_blank(),
              panel.grid.minor = element_blank())+theme(legend.position = "none") 

cat("\n\n...'plot_t-SNE' finished running")
```
```{r plot heat on markers t-SNE}
setwd(paste(getwd(), sep = ""))
dir.create("./output files/")

tsne.by.marker<-as_tibble(tsne.data) %>%
  bind_cols(transformed.chosen.markers)  %>%
  gather(channel, intensity, -V1, -V2) %>%
  mutate(across(channel,factor))%>%
  group_split(channel) %>%
  map(
    ~ggplot(.,aes(x= V1, y= V2, col = intensity)) +
  geom_point(size = 3) +
  scale_color_gradientn(
    colours = colorRampPalette(rev(brewer.pal(n = 11, name = "Spectral")))(5))+
  facet_grid(~ channel, labeller = function(x) label_value(x, multi_line = FALSE)) +
  coord_fixed() +
  theme_bw()+
  theme(strip.text.x = element_text(size = 20),legend.title=element_blank()))%>%
  plot_grid(plotlist = ., align = 'hv', ncol = 8)

png(paste("./output files/",strftime(Sys.time(),"%Y-%m-%d_%H%M%S"),"t-SNE on transformed data.png"),height = 2000,width = 4000)
print(tsne.by.marker)
dev.off()
```

```{r run_UMAP}

# Time ~1 min
set.seed(OVERALL_SEED)
# Run UMAP on all scaled surface markers

# the line below will run UMAP on the data set (to see help page for UMAP, type 
# "?UMAP -- enter" in console)

# you can view UMAP progress by opening up the console below
myumap <-
  umap(transformed.chosen.markers,  # input scaled data
       
       n_neighbors = 15,            # number of nearest neighbors to look at, 
                                    # scales with data set
       
       n_threads = 1,               # this makes UMAP reproducible
       verbose = TRUE)
umap.data = as.data.frame(myumap)

cat("\n\n...'run_UMAP' finished running")

```

```{r plot_UMAP}
# Time <10 sec

# setting aspect ratio for plots
range <- apply(apply(umap.data, 2, range), 2, diff)
graphical.ratio.umap <- (range[1] / range[2])

# UMAP flat dot plot and density dot plot (1 dot = 1 cell)
umap.plot <- data.frame(x = umap.data[, 1], y = umap.data[, 2])

# dot plot
ggplot(umap.plot) + coord_fixed(ratio = graphical.ratio.umap) + 
  geom_point(aes(x = x, y = y), cex = 0.3) + labs(x = "UMAP 1", y = "UMAP 2", 
  title = "PBMC Data on UMAP Axes") + theme_bw() + 
  labs(caption = "Data from Digggins et al., Nat Methods 2017, 14: 275-278 \nFlow Repository: FR-FCM-ZY63") + 
  theme(panel.grid.major = element_blank(),
              panel.grid.minor = element_blank())

# density dot plot
ggplot(umap.plot, aes(x = x, y = y)) + 
  coord_fixed(ratio = graphical.ratio.tsne)+ geom_point(size = 0.5) +geom_density_2d_filled(bins = 39) +scale_fill_manual(values = c("NA","NA","NA","NA","NA","NA","NA","NA","NA","NA","NA","NA","NA",viridis::viridis(28,option = "A"))) +
labs(x = "UMAP 1", y = "UMAP 2", 
  title = "Density on t-SNE Axes") + theme_bw() + 
  labs(caption = "Data from Diggins et al., Nat Methods 2017, 14: 275-278 \nFlow Repository: FR-FCM-ZY63") + 
  theme(panel.grid.major = element_blank(),
              panel.grid.minor = element_blank())+theme(legend.position = "none") 


cat("\n\n...'plot_UMAP' finished running")

```

```{r plot heat on markers UMAP}
setwd(paste(getwd(), sep = ""))
dir.create("./output files/")

umap.by.marker<-as_tibble(umap.data) %>%
  bind_cols(transformed.chosen.markers)  %>%
  gather(channel, intensity, -V1, -V2) %>%
  mutate(across(channel,factor))%>%
  group_split(channel) %>%
  map(
    ~ggplot(.,aes(x= V1, y= V2, col = intensity)) +
  geom_point(size = 3) +
  scale_color_gradientn(
    colours = colorRampPalette(rev(brewer.pal(n = 11, name = "Spectral")))(5))+
  facet_grid(~ channel, labeller = function(x) label_value(x, multi_line = FALSE)) +
  coord_fixed() +
  theme_bw()+
  theme(strip.text.x = element_text(size = 20),legend.title=element_blank()))%>%
  plot_grid(plotlist = ., align = 'hv', ncol = 8)

png(paste("./output files/",strftime(Sys.time(),"%Y-%m-%d_%H%M%S"),"UMAP on transformed data.png"),height = 2000,width = 4000)
print(umap.by.marker)
dev.off()
```

```{r run_FlowSOM_on_t-SNE}
# Time <10 sec

# create flowFrame for FlowSOM input (using t-SNE axes as input)
matrix <- as.matrix(tsne.data)
metadata <-
  data.frame(name = dimnames(matrix)[[2]],
             desc = dimnames(matrix)[[2]])
metadata$range <- apply(apply(matrix, 2, range), 2, diff)
metadata$minRange <- apply(matrix, 2, min)
metadata$maxRange <- apply(matrix, 2, max)
flowframe <- new("flowFrame",
                      exprs = matrix,
                      parameters = AnnotatedDataFrame(metadata))

# implement the FlowSOM on the data by running the line below (to see help page 
# for FlowSOM, type "?FlowSOM --> enter" in console)
fsom <-
  FlowSOM(
    flowframe,      # input flowframe 
    
    colsToUse = c(1:2),  # columns to use 
    
    nClus = CLUSTER_NUM,          # target number of clusters (this can be changed)
    
    seed = OVERALL_SEED  # set seed for reproducibility
  )
FlowSOM.clusters.tsne <-
  GetMetaclusters(fsom)

cat("\n\n...'run_FlowSOM_on_t-SNE' finished running")
```

```{r plot_t-SNE_flowSOM_clusters}
# Time <10 sec
qual_col_pals = brewer.pal.info[brewer.pal.info$category == 'qual',]
col_vector = unlist(mapply(brewer.pal, qual_col_pals$maxcolors, 
                           rownames(qual_col_pals)))

# plot FlowSOM clusters on t-SNE axes
ggplot(tsne.plot) + coord_fixed(ratio=graphical.ratio.tsne) + 
  geom_point(aes(x=x, y=y, color=FlowSOM.clusters.tsne), cex = 0.3) + 
  labs(x = "t-SNE 1", y = "t-SNE 2",title = "FlowSOM Clustering on t-SNE Axes", 
      color = "Cluster") + theme_bw() + 
  guides(colour = guide_legend(override.aes = list(size=4))) +
  scale_color_manual(values = sample(col_vector)) +
  labs(caption = "Data from Diggins et al., Nat Methods 2017, 14: 275-278 \nFlow Repository: FR-FCM-ZY63") + 
  theme(panel.grid.major = element_blank(),
              panel.grid.minor = element_blank())

cat("\n\n...'plot_t-SNE_FlowSOM_clusters' finished running")
```

```{r t-SNE_FlowSOM_varying_cluster_number}
# Time ~ 1-2 min

# run FlowSOM on the t-SNE axes while varying cluster number 
for (i in seq(5,45,by = 10)){

  matrix <- as.matrix(tsne.data)
metadata <-
  data.frame(name = dimnames(matrix)[[2]],
             desc = dimnames(matrix)[[2]])
metadata$range <- apply(apply(matrix, 2, range), 2, diff)
metadata$minRange <- apply(matrix, 2, min)
metadata$maxRange <- apply(matrix, 2, max)
flowframe <- new("flowFrame",
                      exprs = matrix,
                      parameters = AnnotatedDataFrame(metadata))
fsom <-
  FlowSOM(
    flowframe, 
    colsToUse = c(1:2),
    nClus = i,    
    seed = OVERALL_SEED 
  )
FlowSOM.clusters.vary <-
  GetMetaclusters(fsom)

# plot FlowSOM clusters on t-SNE axes
legend.col = round(max(as.numeric(as.vector(FlowSOM.clusters.vary)))/3)
print(ggplot(tsne.plot) + coord_fixed(ratio=graphical.ratio.tsne) + 
  geom_point(aes(x=x, y=y, color=FlowSOM.clusters.vary), cex = 0.3) + 
  labs(x = "t-SNE 1", y = "t-SNE 2",title = "FlowSOM Clustering on t-SNE Axes", 
       color = "Cluster") + theme_bw() + 
  guides(colour = guide_legend(override.aes = list(size=4), 
                               nrow = legend.col)) + 
    scale_color_manual(values = sample(col_vector)) +
  labs(caption = "Data from Diggins et al., Nat Methods 2017, 14: 275-278 \nFlow Repository: FR-FCM-ZY63") + 
  theme(panel.grid.major = element_blank(),
              panel.grid.minor = element_blank()))}

cat("\n\n...'t-SNE_FlowSOM_varying_cluster_number' finished running")
```

```{r FlowSOM_on_original_markers}
# Time ~ 1 min

# create flowFrame for FlowSOM input (using orginal markers as input)
matrix <- as.matrix(transformed.chosen.markers)
metadata <-
  data.frame(name = dimnames(matrix)[[2]],
             desc = dimnames(matrix)[[2]])
metadata$range <- apply(apply(matrix, 2, range), 2, diff)
metadata$minRange <- apply(matrix, 2, min)
metadata$maxRange <- apply(matrix, 2, max)
flowframe <- new("flowFrame",
                      exprs = matrix,
                      parameters = AnnotatedDataFrame(metadata))

# implement the FlowSOM on the data by running the line below 
fsom <-
  FlowSOM(
    flowframe,
    colsToUse = c(1:ncol(transformed.chosen.markers)),
    nClus = CLUSTER_NUM,   
    seed = OVERALL_SEED
  )
FlowSOM.clusters.OG <-
  GetMetaclusters(fsom)

# plot FlowSOM clusters on t-SNE axes
ggplot(tsne.plot) + coord_fixed(ratio=graphical.ratio.tsne) + 
  geom_point(aes(x=x, y=y, color=FlowSOM.clusters.OG), cex = 0.2) + 
  labs(x = "t-SNE 1", y = "t-SNE 2",
       title = "FlowSOM Clustering on Original Markers", color = "Cluster") +
  theme_bw() + scale_color_manual(values = sample(col_vector)) +
  guides(colour = guide_legend(override.aes = list(size=4))) +
  labs(caption = "Data from Diggins et al., Nat Methods 2017, 14: 275-278 \nFlow Repository: FR-FCM-ZY63") + 
  theme(panel.grid.major = element_blank(),
              panel.grid.minor = element_blank())

cat("\n\n...'FlowSOM_on_original_markers' finished running")
```

```{r original_markers_FlowSOM_varying_cluster_number}
# Time ~ 1-2 min

# run FlowSOM on original markers while varying cluster number 
for (i in seq(5,45,by = 10)){

  matrix <- as.matrix(transformed.chosen.markers)
metadata <-
  data.frame(name = dimnames(matrix)[[2]],
             desc = dimnames(matrix)[[2]])
metadata$range <- apply(apply(matrix, 2, range), 2, diff)
metadata$minRange <- apply(matrix, 2, min)
metadata$maxRange <- apply(matrix, 2, max)
flowframe <- new("flowFrame",
                      exprs = matrix,
                      parameters = AnnotatedDataFrame(metadata))
fsom <-
  FlowSOM(
    flowframe, 
    colsToUse = c(1:ncol(transformed.chosen.markers)),
    nClus = i,    
    seed = OVERALL_SEED 
  )
FlowSOM.clusters.vary <-
  GetMetaclusters(fsom)

# plot FlowSOM clusters on t-SNE axes
legend.col = round(max(as.numeric(as.vector(FlowSOM.clusters.vary)))/3)
print(ggplot(tsne.plot) + coord_fixed(ratio=graphical.ratio.tsne) + 
  geom_point(aes(x=x, y=y, color=FlowSOM.clusters.vary), cex = 0.2) + 
  labs(x = "t-SNE 1", y = "t-SNE 2",title = "FlowSOM Clustering on Original Markers", 
       color = "Cluster") + theme_bw() + 
  guides(colour = guide_legend(override.aes = list(size=4), 
                               nrow = legend.col)) + 
    scale_color_manual(values = sample(col_vector)) +
  labs(caption = "Data from Diggins et al., Nat Methods 2017, 14: 275-278 \nFlow Repository: FR-FCM-ZY63") + 
  theme(panel.grid.major = element_blank(),
              panel.grid.minor = element_blank()))}

cat("\n\n...'original_markers_FlowSOM_varying_cluster_number' finished running")
```

```{r FlowSOM_on_UMAP}
# Time <10 sec

# create flowFrame for FlowSOM input (using UMAP axes as input)
matrix <- as.matrix(umap.data)
metadata <-
  data.frame(name = dimnames(matrix)[[2]],
             desc = dimnames(matrix)[[2]])
metadata$range <- apply(apply(matrix, 2, range), 2, diff)
metadata$minRange <- apply(matrix, 2, min)
metadata$maxRange <- apply(matrix, 2, max)
flowframe <- new("flowFrame",
                      exprs = matrix,
                      parameters = AnnotatedDataFrame(metadata))

# implement the FlowSOM on the data by running the line below 
fsom <-
  FlowSOM(
    flowframe,
    colsToUse = c(1:2),
    nClus = CLUSTER_NUM,
    seed = OVERALL_SEED
  )
FlowSOM.clusters.umap <-
  GetMetaclusters(fsom)

# plot FlowSOM clusters on UMAP axes
ggplot(umap.plot) + coord_fixed(ratio=graphical.ratio.umap) + 
  geom_point(aes(x=x, y=y, color=FlowSOM.clusters.umap), cex = 0.3) + 
  labs(x = "UMAP 1", y = "UMAP 2",title = "FlowSOM Clustering on UMAP Axes", 
       color = "Cluster") + theme_bw() + 
  guides(colour = guide_legend(override.aes = list(size=4)))+
  scale_color_manual(values = sample(col_vector))+
  labs(caption = "Data from Diggins et al., Nat Methods 2017, 14: 275-278 \nFlow Repository: FR-FCM-ZY63") + 
  theme(panel.grid.major = element_blank(),
              panel.grid.minor = element_blank())

cat("\n\n...FlowSOM_on_UMAP' finished running")
```

```{r UMAP_FlowSOM_varying_cluster_number}
# Time ~ 1-2 min

# run FlowSOM on the UMAP axes while varying cluster number 
for (i in seq(5,45,by = 10)){

  matrix <- as.matrix(umap.data)
metadata <-
  data.frame(name = dimnames(matrix)[[2]],
             desc = dimnames(matrix)[[2]])
metadata$range <- apply(apply(matrix, 2, range), 2, diff)
metadata$minRange <- apply(matrix, 2, min)
metadata$maxRange <- apply(matrix, 2, max)
flowframe <- new("flowFrame",
                      exprs = matrix,
                      parameters = AnnotatedDataFrame(metadata))
fsom <-
  FlowSOM(
    flowframe, 
    colsToUse = c(1:2),
    nClus = i,    
    seed = OVERALL_SEED 
  )
FlowSOM.clusters.vary <-
  GetMetaclusters(fsom)

# plot FlowSOM clusters on UMAP axes
legend.col = round(max(as.numeric(as.vector(FlowSOM.clusters.vary)))/3)
print(ggplot(umap.plot) + coord_fixed(ratio=graphical.ratio.tsne) + 
  geom_point(aes(x=x, y=y, color=FlowSOM.clusters.vary), cex = 0.3) + 
  labs(x = "UMAP 1", y = "UMAP 2",title = "FlowSOM Clustering on UMAP Axes", 
       color = "Cluster") + theme_bw() + 
  guides(colour = guide_legend(override.aes = list(size=4), 
                               nrow = legend.col)) + 
    scale_color_manual(values = sample(col_vector)) +
  labs(caption = "Data from Diggins et al., Nat Methods 2017, 14: 275-278 \nFlow Repository: FR-FCM-ZY63") + 
  theme(panel.grid.major = element_blank(),
              panel.grid.minor = element_blank()))}

cat("\n\n...'UMAP_FlowSOM_varying_cluster_number' finished running")
```

```{r run_MEM_on_FlowSOM_on_t-SNE}
# Time ~30 sec

# Run MEM on the FlowSOM clusters found from using t-SNE axes
cluster = as.numeric(as.vector((FlowSOM.clusters.tsne)))
MEM.data = cbind(transformed.chosen.markers, cluster)

MEM.values.tf = MEM(
  MEM.data,              # input data (last column must contain cluster values)
  
  transform = FALSE,     # data is already scaled in this case
  cofactor = 1,
  choose.markers = FALSE,
  markers = "all",       # use all transformed, chosen markers from previous 
                         # selection
  
  choose.ref = FALSE,    # reference will be all other cells
  zero.ref = FALSE,
  rename.markers = FALSE,
  new.marker.names = "CD19,CD117,CD11b,CD4,CD8,CD20,CD34,CD61,CD123,CD45RA,CD45,CD10,CD33,CD11c,CD14,CD69,CD15,CD16,CD44,CD38,CD25,CD3,IgM,HLA-DR,CD56", # rename channels for labels
  file.is.clust = FALSE,
  add.fileID = FALSE,
  IQR.thresh = NULL
)

# build MEM heatmap and output enrichment scores
build_heatmaps(
  MEM.values.tf,                # input MEM values
  
  cluster.MEM = "both",         # dendrogram for columns and rows
  
  display.thresh = 2,           # display threshold for MEM scores
  output.files = TRUE,          # makes txt and PDF files for heatmap and MEM 
                                # scores
  
  labels = TRUE,                # include labels in heatmap
  only.MEMheatmap = FALSE
)

cat("\n\n...run_MEM_on_FlowSOM_on_t-SNE' finished running")
```

```{r run_MEM_on_FlowSOM_on_original}
# Time ~30 sec

cluster = as.numeric(as.vector((FlowSOM.clusters.OG)))
MEM.data = cbind(transformed.chosen.markers, cluster)

MEM.values.ogf = MEM(
  MEM.data,       
  transform = FALSE,
  cofactor = 1,
  choose.markers = FALSE,
  markers = "all",  
  choose.ref = FALSE,   
  zero.ref = FALSE,
  rename.markers = FALSE,
  new.marker.names = "CD19,CD117,CD11b,CD4,CD8,CD20,CD34,CD61,CD123,CD45RA,CD45,CD10,CD33,CD11c,CD14,CD69,CD15,CD16,CD44,CD38,CD25,CD3,IgM,HLA-DR,CD56", # rename channels for labels
  file.is.clust = FALSE,
  add.fileID = FALSE,
  IQR.thresh = NULL
)

build_heatmaps(
  MEM.values.ogf,
  cluster.MEM = "both",
  display.thresh = 2,
  output.files = TRUE,
  labels = TRUE,
  only.MEMheatmap = FALSE
)
cat("\n\n...run_MEM_on_FlowSOM_on_original' finished running")
```

```{r run_MEM_on_FlowSOM_on_UMAP}
# Time ~30 sec

cluster = as.numeric(as.vector((FlowSOM.clusters.umap)))
MEM.data = cbind(transformed.chosen.markers, cluster)

MEM.values.uf = MEM(
  MEM.data,   
  transform = FALSE,
  cofactor = 1,
  choose.markers = FALSE,
  markers = "all",  
  choose.ref = FALSE,   
  zero.ref = FALSE,
  rename.markers = FALSE,
  new.marker.names = "CD19,CD117,CD11b,CD4,CD8,CD20,CD34,CD61,CD123,CD45RA,CD45,CD10,CD33,CD11c,CD14,CD69,CD15,CD16,CD44,CD38,CD25,CD3,IgM,HLA-DR,CD56", # rename channels for labels
  file.is.clust = FALSE,
  add.fileID = FALSE,
  IQR.thresh = NULL
)

build_heatmaps(
  MEM.values.uf,
  cluster.MEM = "both", 
  display.thresh = 2,
  output.files = TRUE,
  labels = TRUE,
  only.MEMheatmap = FALSE
)
cat("\n\n...run_MEM_on_FlowSOM_on_UMAP' finished running")
```

```{r run_MEM_on_manually_gated_pops}
# Time ~30 sec

MEM.values.orig = MEM(
  combined.data,    
  transform = TRUE,    
  cofactor = 15,        
  choose.markers = FALSE,
  markers = "12:20,22:23,25:33,35:36,38:40",  
  choose.ref = FALSE,
  zero.ref = FALSE,
  rename.markers = FALSE,
  new.marker.names = "CD19,CD117,CD11b,CD4,CD8,CD20,CD34,CD61,CD123,CD45RA,CD45,CD10,CD33,CD11c,CD14,CD69,CD15,CD16,CD44,CD38,CD25,CD3,IgM,HLA-DR,CD56",
  file.is.clust = FALSE, 
  add.fileID = FALSE,
  IQR.thresh = NULL
)

build_heatmaps(
  MEM.values.orig,      
  cluster.MEM = "both",
  display.thresh = 2,   
  output.files = TRUE,
  labels = TRUE,
  only.MEMheatmap = FALSE
)

cat("\n\n...run_MEM_on_manually_gated_pops' finished running")
```

```{r run_RMSD_all_clusters}
# RMSD to compare labels from all populations (FlowSOM clusters vs. manually
# gated populations)

orig.MEM.scores = as.data.frame(MEM.values.orig[[5]])
rownames(orig.MEM.scores) = paste0(rownames(orig.MEM.scores), " (Manual)")

ogf.MEM.scores = as.data.frame(MEM.values.ogf[[5]])
rownames(ogf.MEM.scores) = paste0(rownames(ogf.MEM.scores), " (OG/fSOM)")

uf.MEM.scores = as.data.frame(MEM.values.uf[[5]])
rownames(uf.MEM.scores) = paste0(rownames(uf.MEM.scores), ' (UMAP/fSOM)')

tf.MEM.scores = as.data.frame(MEM.values.tf[[5]])
rownames(tf.MEM.scores) = paste0(rownames(tf.MEM.scores), ' (t-SNE/fSOM)')

all.MEM.values = as.matrix(rbind(orig.MEM.scores, ogf.MEM.scores, uf.MEM.scores, tf.MEM.scores))

RMSD_vals <-
  MEM_RMSD(
    all.MEM.values,                # input all MEM values from clustering and 
                                   # expert gating
    format = NULL,
    newWindow.heatmaps = FALSE,
    output.matrix = TRUE
  )

cat("\n\n...run_RMSD_on_clusters' finished running")
```
